```{r setoptions, echo = FALSE}
options(scipen = 5, digits = 2)
```

Reproducible Research Programming Assignment I
================================================================================


### About the data

Data from a personal activity monitoring device were collected at 5 minute 
intervals throughout the day. These data consist of the number of steps taken 
during 5 minute intervals each day in October and November 2012 and were collected from
an anonymous individual.

### Loading and preprocessing the data

This codes assumes that the "activity" file has been downloaded and is available
as a csv file in the working directory.

First, the data are read into an R object. The date column is changed into a 
date format and an hour column is created to show the 5-minute intervals in
terms of time of day.

```{r readData}
activity <- read.csv("activity.csv", stringsAsFactors = FALSE)
activity$date <- as.Date(activity$date)
hour <- mapply(function(x, y) paste0(rep(x, y), collapse = ""), 0, 
    4 - nchar(activity$interval))
hour <- paste0(hour, activity$interval)
hour <- format(strptime(hour, format = "%H%M"), format = "%H:%M")
activity$hour <- hour
rm(hour)
```

### What is mean total number of steps taken per day (ignoring missing values)?

The data are first summarized by summing the total per day.

```{r dailySummary}
require(plyr)
dayActivity <- ddply(activity, c("date"), summarise, 
    totalSteps = sum(steps, na.rm = TRUE))
```

A histogram of the total number of steps taken per day is given here.

```{r dailyHist, fig.height = 3, fig.width = 4}
hist(dayActivity$totalSteps, main = "Total number of steps per day",
    xlab = "steps", breaks = 20, cex.main = 0.75)
```

```{r dailyStats}
dayMean <- round(mean(dayActivity$totalSteps),2)
dayMed <- median(dayActivity$totalSteps)
```

The **mean** number of steps per day was `r dayMean`.  
The **median** number of steps per day was `r dayMed`.


### What is the average daily activity pattern?

First, the data are summarised for each 5-minute interval, 
    averaging across all days.

```{r intervalSummary}
intervalActivity <- ddply(activity, c("interval", "hour"), summarise, 
    aveSteps = mean(steps, na.rm = TRUE))
```

A time-series plot of the average number of steps taken per 5-minute interval
is shown here.

```{r intervalPlot, fig.height = 4}
ggplot(intervalActivity, aes(x = hour, y = aveSteps)) +
    geom_line(aes(group = 1)) +
    theme_bw() + 
    scale_x_discrete(breaks = c("00:05", "06:00", "08:35", "12:00", "18:00", "23:55")) +
    xlab("Hour of day") +
    ylab("Average Number of Steps") +
    ggtitle("Average daily activity pattern") +
    geom_vline(x = which(Interval == 0835), linetype = "dashed", color = "red", size = .5)
```

```{r intervalMax}
intervalMax <- intervalActivity$hour[which(intervalActivity$aveSteps ==
    max(intervalActivity$aveSteps))]
```

The 5-minute interval with the maximum number of steps, averaged across all days,
was the interval ending at `r intervalMax`.


### Impute missing values

```{r missingSteps}
stepsMissing <- sum(is.na(activity$steps))
daysMissing <- length(unique(activity$date[is.na(activity$steps)]))
allDays <- length(unique(activity$date))
```

There were `r stepsMissing` 5-minute intervals which were missing values for the
number of steps taken. These data were missing from `r daysMissing` of the 
`r allDays` days in the collection period. During those `r daysMissing` days, 
no data were collected from any interval.

Missing values were imputed with a random sample from the non-missing values in the 
given 5-minute interval. This is a simple and imperfect imputation method, but oh well.


```{r imputeSteps}
set.seed(1353)
imputedActivity <- activity
Interval <- unique(activity$interval)
for(i in seq_along(Interval)) {
    replaceValues <- sample(activity$steps[!is.na(activity$steps) & 
        activity$interval == Interval[i]], 8) 
    imputedActivity$steps[is.na(imputedActivity$steps) & 
        imputedActivity$interval == Interval[i]] <- replaceValues
}
```

After imputing missing values, I recalculated the mean and median of steps per day.

```{r imputeDailyStats}
require(plyr)
imputedDaily <- ddply(imputedActivity, c("date"), summarise, 
    totalSteps = sum(steps))
imputeMean <- round(mean(imputedDaily$totalSteps), 2)
imputeMedian <- median(imputedDaily$totalSteps)
```

A histogram of the dataset including imputed values is plotted below.

```{r imputeHistogram, fig.height = 4, fig.width = 4}
hist(imputedDaily$totalSteps, breaks = 20, xlab = "Total steps per day",
    main = "Missing values imputed by random sampling", cex.main = .75)
```

Imputing values for the missing days results in a distribution that is less
skewed to the right. The new mean and median are larger than for the raw data.
 The new mean is now `r imputeMean` rather than `r dayMean`. The new median
is `r imputeMedian` steps rather than `r dayMed` steps.

There is a difference in means of `r abs(imputeMean - dayMean)` steps, and a
difference in medians of `r abs(imputeMedian - dayMed)` steps.


### Are there differences in activity patterns between weekdays and weekends?

A variable was added to the dataset to indicate whether the activity took place
on a weekday or a weekend. The mean number of steps per interval is then 
calculated separately for weekends and weekends.

```{r weekdays}
imputedActivity$weekday <- weekdays(imputedActivity$date)
imputedActivity$day <- "weekday"
imputedActivity$day[imputedActivity$weekday %in% c("Saturday", "Sunday")] <-
    "weekend"
weekdayActivity <- ddply(imputedActivity, c("day", "interval", "hour"), summarise,
    meanSteps = mean(steps))
```

A visual representation of the mean number of steps per 5-minute interval is given
in the graph below. The steps are plotted against the hour, rather than the 
numeric interval.

```{r plotDaySteps, fig.width = 8, fig.height = 5}
require(ggplot2)
ggplot(weekdayActivity, aes(x = hour, y = meanSteps)) +
    geom_line(aes(group = 1), color = "blue", size = 1) +
    scale_x_discrete(breaks = c("00:05", "06:00", "12:00", "18:00", "23:55")) +
    facet_grid(day ~ .) +
    theme_bw() +
    ggtitle("Daily activity patterns on weekdays versus weekends") +
    xlab("Hour") + ylab("Mean number of steps")
```

Activity appears to begin later in the day and to continue later in the evening on weekends.
The highest mean number of steps on weekdays was higher than on weekends.